\section{Results}\label{results}

\subsection{Training convergence (in-distribution MIMIC validation)}\label{training-convergence-in-distribution-mimic-validation}

Validation top-1 accuracy at convergence reached \textbf{0.992--0.993}
across three training epochs on the held-out MIMIC-IV validation
split. This high in-distribution value reflects the long-tail
concentration of the ACCESS-scope cohort (the top-50 codes cover
the bulk of training rows; see \emph{Cohort and label space}) combined
with effective fine-tuning of a pre-trained biomedical encoder.
Validation loss decreased monotonically across the early-stopping
window, consistent with healthy fine-tuning rather than
memorization.

Aggregate per-epoch metrics (train loss, validation loss,
validation top-1 accuracy) are persisted to
\texttt{models/metrics/train\_*.json} after training completes. Per-record
predictions on MIMIC validation are deliberately not persisted
beyond the in-memory loop --- both to limit checkpoint size and to
preserve the compliance posture that no MIMIC-derived per-record
data leaves the laptop.

\subsection{Out-of-distribution performance (Synthea evaluation cohort)}\label{out-of-distribution-performance-synthea-evaluation-cohort}

The same checkpoint, evaluated on a Synthea-generated cohort of
125 unique synthetic FHIR Conditions (the headline cohort of
\autocite{FungPhantomCodes2026}), yields a substantially lower per-mode
top-1 accuracy in the \textbf{40--50\% range}. Hallucination rate is 0\%
across all four degradation modes by construction --- the
classification head emits one of 50 known ICD-10-CM codes and
cannot fabricate non-existent codes.

The gap between in-distribution validation accuracy (\textasciitilde99\%) and
out-of-distribution Synthea accuracy (\textasciitilde40--50\%) is substantial and
deployment-relevant. Two factors plausibly drive it:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  \textbf{Lexical surface differs systematically.} Synthea's clinical
  text is generated from rule-based modules and differs in vocabulary,
  register, and sentence structure from MIMIC's source notes. The
  classifier learned to map MIMIC-style surface forms to codes;
  Synthea's surface forms are out-of-distribution by construction.
\item
  \textbf{Cohort distribution differs.} The MIMIC training cohort is
  weighted toward conditions that present in critical care; the
  Synthea cohort is weighted toward population-health conditions
  (obesity, prediabetes, hypertension) by virtue of Synthea's
  default module configuration.
\end{enumerate}

This out-of-distribution gap is the most important reported
finding for would-be deployers: a classifier that achieves 99\% on
its source distribution may achieve substantially less on
clinically-realistic but compositionally-different inputs. The
companion paper \autocite{FungPhantomCodes2026} uses this gap as the
non-LLM comparison floor against frontier LLMs evaluated on the
same Synthea inputs.

\subsection{Reproducibility}\label{reproducibility}

A single integer seed (\texttt{seed:\ 42}) is propagated to Python's
\texttt{random}, NumPy, and all PyTorch random number generators (CPU
and MPS) before model construction or data shuffling. Combined
with the deterministic stratified split (also seeded) and full
pinning of dependencies via \texttt{uv} lockfile, this is sufficient to
reproduce training outcomes within floating-point noise on the
same hardware. Bit-exact reproducibility on GPU/MPS additionally
requires \texttt{torch.use\_deterministic\_algorithms(True)}, which incurs
a 10--30\% performance penalty; we do not enable it for v1.

The full set of training hyperparameters is captured in
\texttt{configs/training.yaml} and persisted into each saved checkpoint,
so inference code reads the configuration that produced the
weights rather than relying on a separately-tracked config file.

A credentialed reproducer with PhysioNet access can train a
byte-similar checkpoint by following the recipe documented in the
project repository's \texttt{BENCHMARK.md}. We do not redistribute the
trained weights; reproduction requires running the full pipeline
locally.
