\section{Appendix: Reproduction Recipe}\label{appendix-reproduction-recipe}

This appendix provides everything a credentialed reader needs to
reproduce the trained checkpoint described above from a clean clone
of the project repository. Reproduction requires PhysioNet
credentialing (the MIMIC-IV-FHIR dataset is not redistributed via
this repository).

\subsection{A.1 --- Reproduction sequence}\label{a.1-reproduction-sequence}

\begin{verbatim}
# 1. Install dependencies
uv sync --extra dev

# 2. Download MIMIC-IV-FHIR v2.1 from PhysioNet (requires credentials)
mkdir -p data/mimic/raw && cd data/mimic/raw
wget --user YOUR_PHYSIONET_USERNAME --ask-password \
    https://physionet.org/files/mimic-iv-fhir/2.1/fhir/MimicCondition.ndjson.gz
wget --user YOUR_PHYSIONET_USERNAME --ask-password \
    https://physionet.org/files/mimic-iv-fhir/2.1/fhir/MimicPatient.ndjson.gz
wget --user YOUR_PHYSIONET_USERNAME --ask-password \
    https://physionet.org/files/mimic-iv-fhir/2.1/fhir/MimicEncounter.ndjson.gz
cd ../../..

# 3. Build the parquet derivative splits
uv run phantom-codes prepare \
    --source data/mimic/raw/MimicCondition.ndjson.gz \
    --local-out data/derived

# 4. Fine-tune PubMedBERT on the local data (~15 hr on M1 MPS)
uv run phantom-codes train \
    --config configs/training.yaml \
    --output models/checkpoints/pubmedbert
\end{verbatim}

The full reproduction recipe is documented in the project's
\texttt{BENCHMARK.md} reference file.

\subsection{A.2 --- Pinned dependencies and software versions}\label{a.2-pinned-dependencies-and-software-versions}

A complete \texttt{requirements.txt} produced by \texttt{uv\ export\ -\/-format\ requirements-txt} at the headline-checkpoint commit SHA
is committed alongside this paper for full reproducibility. Pin to
specific package versions so future readers get byte-similar
behavior from the wrapper code.

\subsection{A.3 --- Deterministic configurations}\label{a.3-deterministic-configurations}

The following configuration files fully specify the trained-model
arm:

\begin{itemize}
\tightlist
\item
  \texttt{configs/training.yaml} --- training hyperparameters (peak LR,
  warmup steps, batch size, max sequence length, epochs, patience,
  random seed)
\item
  \texttt{configs/data.yaml} --- data paths, value-set scope filter
\item
  \texttt{src/phantom\_codes/data/access\_valuesets/} --- bundled CMS ACCESS
  Model FHIR ValueSets (v0.9.6) used as the scope filter
\end{itemize}

All three are committed at the checkpoint commit SHA recorded in
the checkpoint manifest sidecar.

\subsection{A.4 --- What's NOT redistributed}\label{a.4-whats-not-redistributed}

Per PhysioNet's responsible-LLM-use policy \autocite{PhysioNet2025},
fine-tuned weights derived from MIMIC are themselves credentialed-
derivative material and are not redistributed via this repository,
HuggingFace Hub, or any cloud bucket. A credentialed reader can
reproduce a byte-similar checkpoint from the recipe in §A.1; we
do not provide a downloadable checkpoint.

The companion paper \autocite{FungPhantomCodes2026} documents the LLM
evaluation arm, which runs entirely on Synthea-generated data and
is fully reproducible without PhysioNet credentialing.
