% Phantom Codes — main paper master file (JAMIA submission).
%
% This is a thin LaTeX skeleton; the actual prose lives in markdown
% files under sections/ and is converted to .tex via pandoc at build
% time (see Makefile). To rebuild the PDF after editing any section:
%
%     cd paper && make phantom_codes
%
% Output: paper/phantom_codes/build/main.pdf (snapshotted to
%         paper/phantom_codes.pdf via `make snapshot-phantom_codes`).
%
% Editing model:
%   - All prose edits happen in sections/*.md (the markdown is canonical)
%   - This main.tex only changes when you add/remove/reorder sections,
%     change document class, add/remove packages, or restyle the title.
%
% When the paper is closer to submission and structure has stabilized,
% the right move is to drop pandoc, convert each section to .tex
% manually (or commit the generated .tex), and switch to a venue-
% specific style file (acl_natbib.sty, neurips.sty, etc.). For now,
% the article class is enough to preview drafts.

\documentclass[11pt]{article}

% ─── Layout & typography ───────────────────────────────────────────────
\usepackage[margin=1in]{geometry}
\usepackage{microtype}                  % subtle kerning improvements
\usepackage{parskip}                    % paragraphs separated by blank lines
                                        % (matches markdown source rhythm)

% ─── Tables, figures, lists ────────────────────────────────────────────
\usepackage{graphicx}
\usepackage{booktabs}                   % \toprule/\midrule/\bottomrule
\usepackage{longtable}                  % multi-page tables (pandoc default)
\usepackage{array}
\usepackage{caption}
\usepackage{calc}                       % needed for pandoc's column-width
                                        % expressions: p{(\columnwidth - X)
                                        % * \real{0.5}}

% ─── Math ──────────────────────────────────────────────────────────────
% amsmath provides \text{}, \frac{}{}, align/equation environments —
% all of which appear in 02_cost_economics.md's metric definitions.
\usepackage{amsmath}
\usepackage{amssymb}                    % \mathbb, \leq, \geq, etc.

% ─── Bibliography ──────────────────────────────────────────────────────
% biblatex with biber backend — modern Unicode-friendly citation engine.
% Pandoc converts [@Key] callouts in markdown sources into \autocite{Key}
% in the generated .tex; biblatex resolves these at build time using
% references.bib and the style configured below.
%
% Citation style options (one-line swap between venues):
%   - vancouver:    superscripted   — biomedical (NEJM AI, JAMIA, BMJ) ← active
%   - numeric-comp: [1, 2]          — IEEE, Nature, NeurIPS
%   - authoryear:   (Devlin, 2019)  — skim-friendly drafting style
%
% Active style: vancouver (AMA-family, superscripted numeric refs).
% NEJM AI's manuscript guidance follows the AMA Manual of Style — this
% setting matches that convention. Switch by changing `style=` only;
% .bib + markdown prose are style-agnostic.
\usepackage[backend=biber, style=vancouver, sortlocale=en_US]{biblatex}
\addbibresource{../references.bib}

% ─── Links & references ────────────────────────────────────────────────
\usepackage[colorlinks=true,
            linkcolor=blue!60!black,
            citecolor=blue!60!black,
            urlcolor=blue!60!black]{hyperref}
\usepackage{xurl}                       % wrap long URLs

% ─── Code & quotes ─────────────────────────────────────────────────────
\usepackage{xcolor}
\usepackage{fancyvrb}                   % code blocks
\usepackage{framed}                     % blockquote backgrounds
\usepackage{csquotes}

% ─── Review-mode niceties (commented; uncomment for NEJM AI submission) ─
% Line numbers throughout the body — most biomedical journals (NEJM AI
% included) want these on the submitted PDF so reviewers can reference
% specific lines. Comment-out for read-throughs; uncomment for submission.
% \usepackage{lineno}
% \linenumbers
% Double-spacing — also a near-universal submission convention.
% \usepackage{setspace}
% \doublespacing

% ─── Pandoc compatibility shims ────────────────────────────────────────
% Pandoc emits these commands in generated .tex; define them so the
% document compiles whether or not pandoc has injected its own preamble.
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\providecommand{\pandocbounded}[1]{#1}

% ─── Title block ───────────────────────────────────────────────────────
\title{%
  Phantom Codes: \\
  \large Hallucination in LLM-Based Clinical Concept Normalization}
% Author block — NEJM AI manuscript convention is name → affiliation → email.
% Update affiliation when institutional placement is finalized; "Independent
% Researcher" is the honest current state. ORCID can be added inline once
% registered.
\author{%
  Brian K. Fung \\[2pt]
  \small Independent Researcher \\[2pt]
  \small\href{mailto:brian@briankfung.com}{\texttt{brian@briankfung.com}}}
\date{\today}

\begin{document}

\maketitle

% Structured abstract — JAMIA convention.
% Required headings: Objective / Materials and Methods / Results /
% Discussion / Conclusion. Word limit: 250 prose words.
\begin{abstract}
\noindent\textbf{Objective.} To characterize how large language
models (LLMs) fail at clinical concept normalization and to separate fabrication of non-existent
codes from abstention, two failure modes typically conflated in
existing benchmarks.

\noindent\textbf{Materials and Methods.} We evaluate eight frontier
LLMs under three prompting modes (zero-shot, constrained, retrieval-
augmented) against a fine-tuned PubMedBERT classifier and string-
matching baselines, on Synthea-generated FHIR Conditions across
four input degradation modes. Each top-1 prediction lands in one of
six outcome buckets that separate fabrication (code does not exist
in CMS-published ICD-10-CM) from abstention (no usable prediction).

\noindent\textbf{Results.} On 125 unique Synthea Conditions $\times$
4 degradation modes, fabrication ranged 0–5.6\% across flagship
configurations and 0–12.8\% across sub-flagship; constrained
prompting eliminated it for every Anthropic and OpenAI model.
Abstention emerged as more consequential for some models —
notably Gemini 2.5 Pro at 88\% empty-prediction under zero-shot
D4. Cost-per-correct ranged 270$\times$; GPT-4o-mini constrained at
\$0.0003 per exact-match was the cost-frontier.

\noindent\textbf{Discussion.} Hallucination appears more
controllable than prior literature suggested at this scale, but
the cohort and abbreviation stress are imperfect proxies for
real-world triggers; larger cohorts with more realistic clinical
text are needed to establish the true effect size.

\noindent\textbf{Conclusion.} Zero-shot LLM prompting is unsafe
for clinical concept normalization; constrained or retrieval-
augmented grounding is required to eliminate fabrication. Among
grounded configurations, cost-per-correct (not accuracy alone)
drives deployment selection: Claude Haiku 4.5 constrained leads at
96.8\% top-1 / \$0.0044 per correct, beating the largest frontier
model (Opus 4.7) at one-third the cost.
\end{abstract}

\tableofcontents
\bigskip

% ─── Body ──────────────────────────────────────────────────────────────
% Each section is generated from sections/*.md via pandoc; see Makefile.
% Add new sections by: (1) creating sections/NN_name.md and
% (2) appending an \input{build/sections/NN_name} line below.

\input{build/sections/00_introduction}
\input{build/sections/01_methodology}
\input{build/sections/02_cost_economics}
\input{build/sections/03_results}
\input{build/sections/04_discussion}
\input{build/sections/05_conclusion}

\bigskip
\hrule
\bigskip

% ─── NEJM AI back-matter ───────────────────────────────────────────────
% Required submission elements per NEJM AI manuscript guidance. Each
% block is a placeholder; fill in before submission. Order follows the
% journal's preferred sequence.

\section*{Funding}
\addcontentsline{toc}{section}{Funding}
This work received no external funding. Compute resources for trained-
model fine-tuning were supplied by the corresponding author's personal
hardware (Apple M1 MacBook Pro). LLM API access (Anthropic, OpenAI,
Google) for the evaluation matrix was funded out of pocket; per-call
costs are itemized in §Cost economics.

\section*{Data availability}
\addcontentsline{toc}{section}{Data availability}
\textbf{MIMIC-IV-FHIR v2.1} is a credentialed dataset distributed by
PhysioNet under their Credentialed Health Data License; access requires
PhysioNet credentialing and CITI training. We do not redistribute MIMIC
data, MIMIC-derivative parquet splits, or model weights fine-tuned on
MIMIC. \textbf{Synthea-generated FHIR Bundles} used for the headline
evaluation are freely redistributable and will be released as
\texttt{benchmarks/synthetic\_v1/} alongside the paper. \textbf{Source
code} (data-preparation pipeline, degradation logic, ICD-10-CM
validator, eval runner, prompts, baselines, trained-model architecture
and training script) is publicly available at the project repository
under the MIT License. Aggregate per-epoch training metrics
(\texttt{models/metrics/*.json}) are available on request; per-record
predictions are not, in keeping with PhysioNet's responsible-LLM-use
policy.

\section*{Code availability}
\addcontentsline{toc}{section}{Code availability}
All source code, configuration files, synthetic test fixtures, and the
manuscript source are publicly available at the project's GitHub
repository under the MIT License. The repository includes the full
evaluation pipeline, allowing any researcher (with or without
PhysioNet credentialing) to reproduce the LLM-evaluation arm against
Synthea-generated inputs.

\section*{Author contributions}
\addcontentsline{toc}{section}{Author contributions}
B.K.F.\ designed the benchmark, implemented the codebase, conducted
the experiments, and wrote the manuscript.

\section*{Competing interests}
\addcontentsline{toc}{section}{Competing interests}
The author declares no competing interests.

\section*{Acknowledgments}
\addcontentsline{toc}{section}{Acknowledgments}
The author gratefully acknowledges the PhysioNet team and the
developers of MIMIC-IV-on-FHIR v2.1 for making this dataset available
to the research community
\autocite{Bennett2024,Bennett2023,Johnson2023,Goldberger2000}.
% Optional personal acknowledgment (uncomment / edit before submission):
% This work was developed during and after a year of full-time
% caregiving for the author's grandmother, whose experience navigating
% clinical care directly informed the methodological emphasis on
% degraded-input robustness throughout this benchmark.

\bigskip
\hrule
\bigskip

% Auto-generated bibliography — biblatex pulls cited entries from
% references.bib, formats per the style configured in the preamble,
% and adds the section to the table of contents via heading=bibintoc.
\printbibliography[heading=bibintoc, title={References}]

\end{document}
