\section{BACKGROUND AND SIGNIFICANCE}\label{background-and-significance}

Medical concept normalization is the entity-linking task of mapping
in-text mentions of biomedical concepts like diagnoses, medications,
labs, and procedures to entries in a standardized ontology or
controlled vocabulary such as ICD-10-CM, LOINC, or RxNorm
\autocite{Miftahutdinov2021,Xu2022}. It sits at the foundation of
healthcare data interoperability. Errors propagate downstream: a
misassigned diagnosis code can shift a hospital's case-mix index,
mis-stratify a research cohort, or corrupt the training data of the
next generation of models. For most of the last decade, the state
of the art for automated coding came from supervised neural
classifiers trained on labeled clinical text
\autocite{Mullenbach2018,Huang2022}. The arrival of general-purpose
large language models has changed the deployment picture: foundation
models can in principle perform code prediction with no task-specific
fine-tuning. Two open questions motivated this work. Do the
fabrication findings reported on earlier model generations still
hold for the frontier models shipping today? And what does the cost
side of the deployment tradeoff look like for the configurations
that do work?

Concept normalization is also a load-bearing component of
\textbf{semantic interoperability}: the property that lets disparate
healthcare information systems exchange data in ways that preserve
clinical meaning, not just syntactic structure. Modern interoperability
stacks (HL7 FHIR, IHE profiles, the OMOP Common Data Model) presume
that values in coded fields resolve to entries in controlled
vocabularies (ICD-10-CM, LOINC, RxNorm, SNOMED CT); they break
silently when those resolutions fail. As clinical data warehouses
scale to support secondary use, multi-site cohort construction, and
FAIR data practices, automated terminology grounding becomes a
critical step in the harmonization pipeline. LLMs are a candidate
mechanism for that step. What determines whether they are viable in
production is not raw accuracy in isolation but the conjunction of
three properties: how often they fabricate codes that do not exist
(silently corrupting downstream harmonization), how often they
abstain (creating tractable gaps a reviewer can fill), and at what
per-prediction cost the resulting pipeline operates at scale.

The literature has documented the fabrication problem but on
generations that are now dated. Soroush et al.\ (2024) reported
that GPT-4 achieved only 46\% exact match on ICD-9 and 34\% on
ICD-10, with a substantial fraction of errors being non-existent
codes rather than real-but-wrong assignments
\autocite{Soroush2024}. Lee and Lindsey documented that LLMs
systematically struggle to distinguish real codes from
plausibly-formatted fakes \autocite{Lee2024}. Both findings are an
instance of the broader medical-hallucination phenomenon
\autocite{Kim2025,Ji2023}, and both predate Anthropic's Opus 4.7,
OpenAI's GPT-5.5, and Google's Gemini 3 Flash Preview. Whether the
fabrication rates reported on GPT-4 generalize to the current
generation has not been re-measured.

Cost is largely absent from the published evaluations. The field's
response to hallucination has been architectural: retrieval-grounded
pipelines \autocite{DasBaksi2024,Sarvari2025}, agentic walkers
\autocite{Motzfeldt2025}, neuro-symbolic verifiers
\autocite{HybridCode2025}, and broader benchmark expansions
\autocite{Almeida2025,Singh2025,Li2025,Gershon2025}. Across this
body of work, hallucination is rarely treated as an explicit
outcome, and per-prediction cost is rarely reported at all. As
LLM-augmented coding moves toward production deployment, the
operationally relevant question is which (model, prompting mode)
combination achieves a useful accuracy at a tolerable cost. The
literature does not yet answer that question.

We present \textbf{Phantom Codes}, whose primary contribution is a
reproducible evaluation framework for LLM-based terminology
grounding, demonstrated on the most recent frontier-model
generation. The framework has four reusable components. \emph{First},
a \textbf{six-way outcome taxonomy} in which fabrication of
non-existent codes (hallucination) and abstention
(\texttt{no\_prediction}) are explicit, mutually exclusive buckets;
the taxonomy generalizes to any closed-vocabulary grounding task
including LOINC and RxNorm. \emph{Second}, a \textbf{within-model
ablation} across zero-shot, constrained, and retrieval-augmented
prompting modes that holds the model and input fixed and varies
only whether a candidate list is provided, isolating ``wandering
off the menu'' from genuine ignorance. \emph{Third}, an
\textbf{abbreviation-stress mode (D4)} that replaces canonical
display strings with clinical jargon (T2DM, HTN, CKD-3) to expose
where string-matching baselines collapse and where genuine semantic
mapping is required; D4 functions as a robustness probe for any
LLM-grounding pipeline. \emph{Fourth}, \textbf{per-prediction cost
as a first-class outcome} alongside accuracy; we report cost-per-
correct (USD per exact-match outcome) as the deployment-relevant
normalization, computed from per-call token usage against published
per-token rates.

The empirical application of this framework spans 24 (model,
prompting-mode) configurations across Claude Opus 4.7, Sonnet 4.6,
and Haiku 4.5; GPT-5.5 and GPT-4o-mini; and Gemini 2.5 Pro, 2.5
Flash, and 3 Flash Preview --- several released within weeks of the
evaluation date. The headline matrix runs against Synthea-generated
FHIR Bundles \autocite{Walonoski2018}, a freely-redistributable
synthetic patient dataset that gives full reproducibility from a
pinned Synthea version and seed.
