% PubMedBERT fine-tuning — companion technical report (planned arXiv preprint).
%
% Standalone tech report describing the trained PubMedBERT
% classification arm in full. Originally appeared as Supplementary §S3
% of the Phantom Codes JAMIA paper (paper/phantom_codes/main.tex);
% lifted into its own preprint so the JAMIA paper can focus on the LLM
% hallucination evaluation and the methodology has room to be
% presented properly.
%
% Build:
%     cd paper && make pubmedbert
%
% Output: paper/pubmedbert/build/main.pdf (snapshotted to
%         paper/pubmedbert_finetuning.pdf via `make snapshot-pubmedbert`).
%
% Editing model:
%   - Prose edits happen in sections/*.tex.
%   - This file only changes when sections are added/removed/reordered.

\documentclass[11pt]{article}

% ─── Layout & typography ───────────────────────────────────────────────
\usepackage[margin=1in]{geometry}
\usepackage{microtype}
\usepackage{parskip}

% ─── Tables, figures, lists ────────────────────────────────────────────
\usepackage{graphicx}
\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{caption}
\usepackage{calc}

% ─── Math ──────────────────────────────────────────────────────────────
\usepackage{amsmath}
\usepackage{amssymb}

% ─── Bibliography ──────────────────────────────────────────────────────
% Shared references.bib with main.tex / supplementary.tex; same Vancouver
% style so cross-references render consistently across the three documents.
\usepackage[backend=biber, style=vancouver, sortlocale=en_US]{biblatex}
\addbibresource{../references.bib}

% ─── Links & references ────────────────────────────────────────────────
\usepackage[colorlinks=true,
            linkcolor=blue!60!black,
            citecolor=blue!60!black,
            urlcolor=blue!60!black]{hyperref}
\usepackage{xurl}

% ─── Code & quotes ─────────────────────────────────────────────────────
\usepackage{xcolor}
\usepackage{fancyvrb}
\usepackage{framed}
\usepackage{csquotes}

% ─── Pandoc legacy shim ────────────────────────────────────────────────
% \tightlist still appears in section sources (suppresses inter-item
% spacing in lists). Define it so the document compiles standalone.
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}

% ─── Title block ───────────────────────────────────────────────────────
\title{%
  Local Fine-Tuning of PubMedBERT for ACCESS-Scope ICD-10-CM
  Classification under PhysioNet Compliance Constraints \\
  \large A Companion Technical Report to \emph{Phantom Codes}}
\author{%
  Brian K. Fung \\[2pt]
  \small Independent Researcher \\[2pt]
  \small\href{mailto:brian@briankfung.com}{\texttt{brian@briankfung.com}}}
\date{\today}

\begin{document}

\maketitle

\begin{abstract}
\noindent We document a fully-local fine-tuning pipeline for
PubMedBERT-base on MIMIC-IV-FHIR v2.1 restricted to ICD-10-CM codes
in the CMS ACCESS Model FHIR Implementation Guide. Every step from
data loading through training through checkpoint persistence runs on
consumer Apple Silicon (M1 MacBook Pro, 16 GB unified memory) via
PyTorch's MPS backend, with no MIMIC content uploaded to any cloud
service, training platform, or telemetry endpoint. The classifier
attains 0.992--0.993 validation top-1 accuracy on the held-out
MIMIC-IV split and 40--50\% top-1 on a Synthea-generated
out-of-distribution evaluation cohort, with 0\% hallucination by
construction (the head emits one of 50 known ICD-10-CM codes). This
trained checkpoint serves as the non-LLM baseline in the companion
paper \emph{Phantom Codes} \autocite{FungPhantomCodes2026}, which
evaluates frontier LLMs on the same Synthea cohort under a 6-way
hallucination-aware outcome taxonomy. The contribution of this
report is methodological: research-grade clinical-classifier
fine-tuning is reachable on consumer hardware, the data-residency
story is dramatically simpler than cloud GPU paths, and the same
code path supports CUDA backends if cloud-compute authorization is
in place.
\end{abstract}

\tableofcontents
\bigskip

% ─── Body ──────────────────────────────────────────────────────────────
\input{sections/00_introduction}
\input{sections/01_data_and_compliance}
\input{sections/02_architecture}
\input{sections/03_hardware_and_optimization}
\input{sections/04_results}
\input{sections/05_discussion}
\input{sections/06_appendix}

\bigskip
\hrule
\bigskip

% ─── Back-matter ───────────────────────────────────────────────────────
\section*{Funding}
\addcontentsline{toc}{section}{Funding}
This work received no external funding. Compute resources for
trained-model fine-tuning were supplied by the corresponding
author's personal hardware (Apple M1 MacBook Pro).

\section*{Data availability}
\addcontentsline{toc}{section}{Data availability}
\textbf{MIMIC-IV-FHIR v2.1} is a credentialed dataset distributed by
PhysioNet under their Credentialed Health Data License; access
requires PhysioNet credentialing and CITI training. We do not
redistribute MIMIC data, MIMIC-derivative parquet splits, or model
weights fine-tuned on MIMIC. \textbf{Source code} (data-preparation
pipeline, training script, configuration files) is publicly
available at the project repository under the MIT License.
Aggregate per-epoch training metrics
(\texttt{models/metrics/*.json}) are available on request;
per-record predictions are not, in keeping with PhysioNet's
responsible-LLM-use policy.

\section*{Code availability}
\addcontentsline{toc}{section}{Code availability}
All source code, configuration files, and the manuscript source are
publicly available at the project's GitHub repository under the MIT
License. The repository includes the full training pipeline; a
credentialed reader can train a byte-similar checkpoint by
following the recipe in §A.1.

\section*{Author contributions}
\addcontentsline{toc}{section}{Author contributions}
B.K.F.\ designed the pipeline, implemented the codebase, conducted
the training run, and wrote the manuscript.

\section*{Competing interests}
\addcontentsline{toc}{section}{Competing interests}
The author declares no competing interests.

\section*{Acknowledgments}
\addcontentsline{toc}{section}{Acknowledgments}
The author gratefully acknowledges the PhysioNet team and the
developers of MIMIC-IV-on-FHIR v2.1 for making this dataset
available to the research community
\autocite{Bennett2024,Bennett2023,Johnson2023,Goldberger2000}.

\bigskip
\hrule
\bigskip

\printbibliography[heading=bibintoc, title={References}]

\end{document}
