% Phantom Codes — paper master file.
%
% This is a thin LaTeX skeleton; the actual prose lives in markdown
% files under sections/ and is converted to .tex via pandoc at build
% time (see Makefile). To rebuild the PDF after editing any section:
%
%     cd paper && make pdf
%
% Output: paper/build/paper.pdf
%
% Editing model:
%   - All prose edits happen in sections/*.md (the markdown is canonical)
%   - This main.tex only changes when you add/remove/reorder sections,
%     change document class, add/remove packages, or restyle the title.
%
% When the paper is closer to submission and structure has stabilized,
% the right move is to drop pandoc, convert each section to .tex
% manually (or commit the generated .tex), and switch to a venue-
% specific style file (acl_natbib.sty, neurips.sty, etc.). For now,
% the article class is enough to preview drafts.

\documentclass[11pt]{article}

% ─── Layout & typography ───────────────────────────────────────────────
\usepackage[margin=1in]{geometry}
\usepackage{microtype}                  % subtle kerning improvements
\usepackage{parskip}                    % paragraphs separated by blank lines
                                        % (matches markdown source rhythm)

% ─── Tables, figures, lists ────────────────────────────────────────────
\usepackage{graphicx}
\usepackage{booktabs}                   % \toprule/\midrule/\bottomrule
\usepackage{longtable}                  % multi-page tables (pandoc default)
\usepackage{array}
\usepackage{caption}
\usepackage{calc}                       % needed for pandoc's column-width
                                        % expressions: p{(\columnwidth - X)
                                        % * \real{0.5}}

% ─── Math ──────────────────────────────────────────────────────────────
% amsmath provides \text{}, \frac{}{}, align/equation environments —
% all of which appear in 02_cost_economics.md's metric definitions.
\usepackage{amsmath}
\usepackage{amssymb}                    % \mathbb, \leq, \geq, etc.

% ─── Bibliography ──────────────────────────────────────────────────────
% biblatex with biber backend — modern Unicode-friendly citation engine.
% Pandoc converts [@Key] callouts in markdown sources into \autocite{Key}
% in the generated .tex; biblatex resolves these at build time using
% references.bib and the style configured below.
%
% Citation style options (one-line swap between venues):
%   - vancouver:    superscripted   — biomedical (NEJM AI, JAMIA, BMJ) ← active
%   - numeric-comp: [1, 2]          — IEEE, Nature, NeurIPS
%   - authoryear:   (Devlin, 2019)  — skim-friendly drafting style
%
% Active style: vancouver (AMA-family, superscripted numeric refs).
% NEJM AI's manuscript guidance follows the AMA Manual of Style — this
% setting matches that convention. Switch by changing `style=` only;
% .bib + markdown prose are style-agnostic.
\usepackage[backend=biber, style=vancouver, sortlocale=en_US]{biblatex}
\addbibresource{references.bib}

% ─── Links & references ────────────────────────────────────────────────
\usepackage[colorlinks=true,
            linkcolor=blue!60!black,
            citecolor=blue!60!black,
            urlcolor=blue!60!black]{hyperref}
\usepackage{xurl}                       % wrap long URLs

% ─── Code & quotes ─────────────────────────────────────────────────────
\usepackage{xcolor}
\usepackage{fancyvrb}                   % code blocks
\usepackage{framed}                     % blockquote backgrounds
\usepackage{csquotes}

% ─── Review-mode niceties (commented; uncomment for NEJM AI submission) ─
% Line numbers throughout the body — most biomedical journals (NEJM AI
% included) want these on the submitted PDF so reviewers can reference
% specific lines. Comment-out for read-throughs; uncomment for submission.
% \usepackage{lineno}
% \linenumbers
% Double-spacing — also a near-universal submission convention.
% \usepackage{setspace}
% \doublespacing

% ─── Pandoc compatibility shims ────────────────────────────────────────
% Pandoc emits these commands in generated .tex; define them so the
% document compiles whether or not pandoc has injected its own preamble.
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\providecommand{\pandocbounded}[1]{#1}

% ─── Title block ───────────────────────────────────────────────────────
\title{%
  Phantom Codes: \\
  \large Hallucination in LLM-Based Clinical Concept Normalization}
% Author block — NEJM AI manuscript convention is name → affiliation → email.
% Update affiliation when institutional placement is finalized; "Independent
% Researcher" is the honest current state. ORCID can be added inline once
% registered.
\author{%
  Brian K. Fung \\[2pt]
  \small Independent Researcher \\[2pt]
  \small\href{mailto:brian@briankfung.com}{\texttt{brian@briankfung.com}}}
\date{\today}

\begin{document}

\maketitle

% Structured abstract — NEJM AI / AMA convention.
% Target word count: 250-300 words across all four sections combined.
% Numbers in *Results* stay TBD until the headline run completes; rest
% of the abstract is settled (matches the introduction + methods).
\begin{abstract}
\noindent\textbf{Background.} Large language models can perform
clinical concept normalization (mapping unstructured clinical text to
controlled-vocabulary identifiers such as ICD-10-CM) without
task-specific fine-tuning, but they may fabricate codes that do not
exist in the target vocabulary. Existing benchmarks treat such
fabrications as ordinary errors, conflating two distinct failure
modes (real-but-wrong code vs.\ non-existent code) and obscuring the
mechanism by which constrained-decoding mitigations help.

\noindent\textbf{Methods.} We introduce \emph{Phantom Codes}, a
reproducible benchmark that evaluates frontier LLMs (Claude Opus 4.7,
Sonnet 4.6, Haiku 4.5; GPT-5.5 and a GPT-4-class economy comparator;
Gemini 2.5 Pro and Flash) under three prompting modes (zero-shot,
constrained, retrieval-augmented) against a fine-tuned PubMedBERT
classifier head, a sentence-transformer retrieval baseline, and three
constrained-only string-matching baselines. Inputs span four
degradation modes ranging from full FHIR Conditions to
abbreviation-only one-sentence summaries. Every prediction is
assigned to one of five mutually exclusive outcome buckets:
\emph{exact match}, \emph{category match}, \emph{chapter match},
\emph{out of domain}, and \emph{hallucination} (predicted code does
not exist in CMS-published ICD-10-CM). Trained models are fine-tuned
locally on credentialed MIMIC-IV-FHIR data; the evaluation matrix
runs entirely against Synthea-generated synthetic FHIR Bundles,
ensuring full compliance with PhysioNet's responsible-LLM-use policy
and full reproducibility without credentialed-data access.

\noindent\textbf{Results.} [TBD — populated from headline run on
Synthea cohort. Will report per-model and per-mode outcome
distributions across all five buckets, with hallucination rates as
the primary outcome and cost-per-correct-prediction as a secondary
deployment-relevance metric.]

\noindent\textbf{Conclusions.} [TBD — three sentences once results
are in; current expectation is that hallucination rates differ
substantially across prompting modes within the same model, that
constrained decoding reduces but does not eliminate hallucination,
and that the trained classifier head sets a clean comparison floor
on a closed vocabulary.]
\end{abstract}

\tableofcontents
\bigskip

% ─── Body ──────────────────────────────────────────────────────────────
% Each section is generated from sections/*.md via pandoc; see Makefile.
% Add new sections by: (1) creating sections/NN_name.md and
% (2) appending an \input{build/sections/NN_name} line below.

\input{build/sections/00_introduction}
\input{build/sections/01_methodology}
\input{build/sections/02_cost_economics}
\input{build/sections/03_results}
\input{build/sections/04_discussion}
\input{build/sections/05_conclusion}

\bigskip
\hrule
\bigskip

% ─── NEJM AI back-matter ───────────────────────────────────────────────
% Required submission elements per NEJM AI manuscript guidance. Each
% block is a placeholder; fill in before submission. Order follows the
% journal's preferred sequence.

\section*{Funding}
\addcontentsline{toc}{section}{Funding}
This work received no external funding. Compute resources for trained-
model fine-tuning were supplied by the corresponding author's personal
hardware (Apple M1 MacBook Pro). LLM API access (Anthropic, OpenAI,
Google) for the evaluation matrix was funded out of pocket; per-call
costs are itemized in §Cost economics.

\section*{Data availability}
\addcontentsline{toc}{section}{Data availability}
\textbf{MIMIC-IV-FHIR v2.1} is a credentialed dataset distributed by
PhysioNet under their Credentialed Health Data License; access requires
PhysioNet credentialing and CITI training. We do not redistribute MIMIC
data, MIMIC-derivative parquet splits, or model weights fine-tuned on
MIMIC. \textbf{Synthea-generated FHIR Bundles} used for the headline
evaluation are freely redistributable and will be released as
\texttt{benchmarks/synthetic\_v1/} alongside the paper. \textbf{Source
code} (data-preparation pipeline, degradation logic, ICD-10-CM
validator, eval runner, prompts, baselines, trained-model architecture
and training script) is publicly available at the project repository
under the MIT License. Aggregate per-epoch training metrics
(\texttt{models/metrics/*.json}) are available on request; per-record
predictions are not, in keeping with PhysioNet's responsible-LLM-use
policy.

\section*{Code availability}
\addcontentsline{toc}{section}{Code availability}
All source code, configuration files, synthetic test fixtures, and the
manuscript source are publicly available at the project's GitHub
repository under the MIT License. The repository includes the full
evaluation pipeline, allowing any researcher (with or without
PhysioNet credentialing) to reproduce the LLM-evaluation arm against
Synthea-generated inputs.

\section*{Author contributions}
\addcontentsline{toc}{section}{Author contributions}
B.K.F.\ designed the benchmark, implemented the codebase, conducted
the experiments, and wrote the manuscript.

\section*{Competing interests}
\addcontentsline{toc}{section}{Competing interests}
The author declares no competing interests.

\section*{Acknowledgments}
\addcontentsline{toc}{section}{Acknowledgments}
[Optional — populate before submission if appropriate.]

\bigskip
\hrule
\bigskip

% Auto-generated bibliography — biblatex pulls cited entries from
% references.bib, formats per the style configured in the preamble,
% and adds the section to the table of contents via heading=bibintoc.
\printbibliography[heading=bibintoc, title={References}]

\end{document}
